Topological Equivariant Artist Model

### 2021-03-24 slides/main.tex
still too many slides probably

\usepackage{tabulary}
\usepackage{booktabs}
\usepackage{multirow}

@@

\begin{frame}{Visualizations are tuned to continuinity\cite{toryRethinkingVisualizationHighlevel2004}}
    \begin{figure}
        \includegraphics[height=.5\textheight]{figures/intro/dataset_diagram.png}
        \caption{Data Representation, MayaVi 4.7.2 docs\cite{DataRepresentationMayavi}}
    \end{figure}
    \begin{enumerate}
        \item Matplotlib\cite{hunterMatplotlib2DGraphics2007}, 
        \item D3 \cite{bostockDataDrivenDocuments2011}
        \item VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}, MayaVi\cite{RamachandranMayaVI2011}, ParaView\cite{ahrens2005paraview}, Titan\cite{brianwylieUnifiedToolkitInformation2009}
    \end{enumerate}
\end{frame}

@@

\begin{frame}{Encode variable types in a schema like fiber \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}}
    \begin{figure}[H]
        \centering
        \includegraphics[height=.5\textwidth]{figures/math/fiber.png}
        \label{fig:data_fiber_example}
    \end{figure}
    \begin{block}{Fiber components can be acted on independently}
        \begin{equation}
            \dfiber = \dfiber_0 \times \ldots \times \dfiber_i \times \ldots \times \dfiber_n
        \end{equation}
    \end{block}
\end{frame}

\begin{frame}{}
    \begin{figure}[H]
        \includegraphics[width=1\textwidth]{figures/math/base.png}
        \caption{}
        \label{fig:base_example}
    \end{figure}
    \begin{block}{\dbase\ is an indexing space regardless of components}
        \begin{equation}
            \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
        \end{equation}
    \end{block}
\end{frame}

@@

\subsection{Graphics}
\begin{frame}{Graphic Bundle}
    The graphics bundle is a tuple $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$ defined by the projection map $\pi$
    \begin{equation}
        \begin{tikzcd}[ampersand replacement=\&]
            \gfiber \arrow[r, hook] \& \gtotal \arrow[d, "\pi"'] \\
                              \& \gbase \arrow[u, "\gsection"', bend right]
        \end{tikzcd}
    \end{equation}
    where \gsection\ is the fully encoded graphic. 
    \pause
    \begin{block}{Example: 2D opaque image}
    The target display is $\gfiber=\reals^5$ with elements 
    \begin{equation*}
    (x,\, y,\, r,\, g,\, b) \in \gfiber
    \end{equation*}
    returned by \gsection\ such that a graphic has color and 2D position.
    \end{block}

\end{frame}
\begin{frame}{Graphic Continuity}
    The surjective map $\vindex: \gbase \rightarrow \dbase$
    \begin{equation}
        \begin{tikzcd}[ampersand replacement=\&]
            \dtotal \arrow[d, "\pi"'] \& \gtotal \arrow[d, "\pi"'] \\
            \dbase                   \& \gbase \arrow[l, "\vindex"']
        \end{tikzcd}
    \end{equation}
    goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\dbasepoint$ in data space. 
\pause
    \begin{figure}[H]
        \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
        \label{fig:graphic_retraction_map}
    \end{figure}
\end{frame}

\subsection{Artist}
\begin{frame}{Topological Equivariant Artist Model}
The topological artist $\vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)$ is sheaf map
\begin{equation}
    \label{eq:artist_diagram}
    \begin{tikzcd}[ampersand replacement=\&]
        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] \& \vtotal \arrow[d, "\pi"] \& \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] \& \gtotal \arrow[ld, "\pi"] \\
                                              \& \dbase                  \& \gbase \arrow[l, "\vindex"']                                              \&                    
        \end{tikzcd}
\end{equation}
that carries homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019} 
\begin{equation}\huge
\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
\end{equation} 
\end{frame}

\begin{frame}{Existing models don't have continuity}
        \begin{description}
            \item[language model] APT, GoG: syntax, semantics, and grammar of graphics (Mackinlay, Wilkenson  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005})
            \item[functional dependencies] constrained maps between data and visual representation(Sugibuchi \cite{sugibuchiFramwork2009}) 
            \item[category theory] data, representation, and understanding are commutative (Vickers \cite{vickersUnderstandingViz2013})
            \item[algebraic process] data ($\alpha$) and viz ($\omega$) transforms are symmetric (Kindlmann and Scheidegger \cite{kindlmannAlgebraicProcessVisualization2014})
            \begin{columns}
                \column{.5\textwidth}
                \begin{description}
                    \item[\vtotal] visual components
                    \item[$\alpha$] \monoid
                    \item[$\omega$] $\varphi(\monoid)$
                \end{description}
                \column{.5\textwidth}
                \begin{equation*}
                    \begin{tikzcd}[ampersand replacement=\&]
                        \dtotal \arrow[d, "\alpha"'] \arrow[r, "r_1"] \& \vtotal \arrow[r, "\nu"]  \& \gtotal \arrow[d, "\omega"] \\
                        \dtotal \arrow[r, "r_2"']                     \& \vtotal \arrow[r, "\nu"'] \& \gtotal                  
                    \end{tikzcd}
                    \end{equation*}

            \end{columns}

@@

\begin{frame}{Visual Bundle}
    The visual bundle is a tuple $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ defined by the projection map $\pi$
    \begin{equation}
        \begin{tikzcd}[ampersand replacement=\&]
            \vfiber \arrow[r, hook] \& \vtotal \arrow[d, "\pi"'] \\
                              \& \dbase \arrow[u, "\vsection"', bend right]
        \end{tikzcd}
    \end{equation}
    where \vsection\ is the visual variable encoding\cite{bertinIIPropertiesGraphic2011} of the data section \dsection. 
    \begin{block}{Example: position and color}
        Given an artist with parameters $\{xpos, ypos, color\}$, a sample visual section \vsection\ could be $\{.5, .5, (255, 20,147)\}$
    \end{block}
\end{frame}

@@

\begin{frame}{Partial Order Equivariance and Stevens Scales \cite{stevensTheoryScalesMeasurement1946}}
    \begin{figure}
        \includegraphics[width=1\textwidth]{figures/math/monoid_equivariant.png}
    \end{figure}

\begin{table}[H]
    \begin{tabulary}{\textwidth}{llL}
        nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
        ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
        interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
        ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $
    \end{tabulary}
\end{table}
\end{frame}

@@

\begin{frame}{Assembly Function Factory}
\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/math/q_hat.png}
\end{figure}
\begin{equation}\label{eq:qhat_q_s}
    \vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
\end{equation} 
such \gbasepoint\ can be factored out when $\vpreimg = \gbasepoint$
\end{frame}

@@

    def __call__(self, value):
        # convert value to a color
        return [mcolors.to_rgba(self._mapping[v]) for v in values]
\end{minted}
\pause
That we can test for action equivariance
\begin{minted}{python}
def test_nominal(values, encoder):
    m1 = list(zip(values, encoder(values)))
    random.shuffle(values)
    m2 = list(zip(values, encoder(values)))
    assert sorted(m1) == sorted(m2)
\end{minted}
\end{frame}
\subsection{Data}
\begin{frame}[fragile]{Fiber Bundle}
\begin{minted}{python}
@dataclass
class FiberBundle:
"""
Attributes
----------
K: {'tables': []}
F: {variable name: type}
"""
    K: dict 
    F: dict
\end{minted}
\end{frame}

@@

\section{References}
\begin{frame}[allowframebreaks]{References}
\printbibliography
\end{frame}
\appendix 
\begin{frame}{Fiber is all possible values a variable can be \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}}
    Given a space of all possible values \ftotal\
    \begin{equation}
        \label{eq:data_types}
        \begin{tikzcd}[ampersand replacement=\&]
            \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] \& \ftotal \arrow[d, "\pi"] \\
            \fnames \arrow[r, "\fsection"']                  \& \ftypes       
        \end{tikzcd}
    \end{equation}
    a fiber component is the restricted space $\ftotal_{\fsection(\fname)}$. 
    \begin{equation}
        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
    \end{equation}
    \begin{description}
        \item[\ftypes] data types of the variables in the dataset 
        \item[\ftotal] disjoint union of all values of type $\ftype \in \ftypes$ 
        \item[\fnames] variable names, $\fname \in \fnames$
        \item[\fttype] \ftotal\ restricted to the data type of a named variable   
    \end{description}
\end{frame}
\begin{frame}{Monoid actions}
    A monoid \monoid\ is a set with
    \begin{description}
        \item[associative binary operator] $\ast:\monoid \times \monoid\rightarrow \monoid$
        \item[identity element] $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. 
    \end{description}
    \pause
    \begin{block}{left monoid action}
    A set \dfiber\ with an action\cite{nlab:action} $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
        \begin{align*}
            \textbf{associativity}\;& \text{for all } f,g \in \monoid \text{ and } x\in \dfiber,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
            \textbf{identity}\;& \text{for all } x\in \dfiber, e\in \monoid,\,  e\bullet x = x 
        \end{align*}
    \end{block}
\end{frame}
\begin{frame}{Keeping track of sections with sheafs}
    Restriction maps of a sheaf describe how local $\iota^*\dsection$ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.

>>>>

\usepackage{tabulary}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{multirow}

@@

\begin{frame}{Visualizations are tuned to continuinity\cite{toryRethinkingVisualizationHighlevel2004}}
    \begin{figure}
        \includegraphics[height=.5\textheight]{figures/intro/dataset_diagram.png}
        \caption{Data Representation, MayaVi 4.7.2 docs\cite{DataRepresentationMayavi}}
    \end{figure}
    \begin{enumerate}
        \item Matplotlib\cite{hunterMatplotlib2DGraphics2007} $\rightarrow$Seaborn\cite{waskom2020seaborn}, xarray \cite{hoyer2017xarray} 
        \item D3 \cite{bostockDataDrivenDocuments2011}
        \item VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}(MayaVi\cite{RamachandranMayaVI2011})$\rightarrow$ Titan\cite{brianwylieUnifiedToolkitInformation2009}, ParaView\cite{ahrens2005paraview}
    \end{enumerate}
\end{frame}

@@

\begin{frame}{Encode variable types in a schema like fiber \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}}
    \begin{figure}[H]
        \centering
        \includegraphics[height=.3\textwidth]{figures/math/fiber.png}
        \label{fig:data_fiber_example}
    \end{figure}
    \begin{block}{Monoid actions define the structure on $\dfiber_i$}
        \begin{equation}
            \dfiber = \dfiber_0 \times \ldots \times \dfiber_i \times \ldots \times \dfiber_n
        \end{equation}

        $\monoid_{i}$ actions (e.g. rotation, partial ordering) 
        \begin{equation}
            \bullet: \monoid_i\times \dfiber_i \rightarrow \dfiber_i
        \end{equation}
        must be associative and have an identity element.

    \end{block}
\end{frame}

\begin{frame}{}
    \begin{figure}[H]
        \includegraphics[width=1\textwidth]{figures/math/base.png}
        \label{fig:base_example}
    \end{figure}
    \begin{block}{\dbase\ is an indexing space regardless of components}
        \begin{equation}
            \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
        \end{equation}
    \end{block}
\end{frame}

@@

\subsection{Graphics}
\begin{frame}{Graphic Bundle $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$}
    Continuity is preserved via the deformation retraction $\vindex: \gbase\rightarrow\dbase$
    \begin{equation}
        \begin{tikzcd}[ampersand replacement=\&]
            F \arrow[r, hook] \& E \arrow[d, "\pi"']              \& D \arrow[r, hook] \& H \arrow[d, "\pi"']                                 \\
                              \& K \arrow[u, "\tau"', bend right] \&                   \& S \arrow[ll, "\xi"'] \arrow[u, "\rho"', bend right]
            \end{tikzcd}
        \end{equation}
    \gfiber\ is a proxy for the target display, for example $(x,\, y,\, r,\, g,\, b) \in \gfiber$
    \begin{figure}[H]
        \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
        \label{fig:graphic_retraction_map}
    \end{figure}
\end{frame}

\subsection{Artist}
\begin{frame}{Topological Equivariant Artist Model}
The topological artist is a sheaf map
\begin{equation}
\vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
\end{equation}
that carries homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019} 
\begin{equation}
\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
\end{equation} 
and is implemented via an intermediate visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$
\begin{equation}
    \label{eq:artist_diagram}
    \begin{tikzcd}[ampersand replacement=\&]
        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] \& \vtotal \arrow[d, "\pi"] \& \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] \& \gtotal \arrow[ld, "\pi"] \\
                                              \& \dbase                  \& \gbase \arrow[l, "\vindex"']                                              \&                    
        \end{tikzcd}
\end{equation}
\end{frame}

\begin{frame}{Existing models don't have continuity, many evaluative}
        \begin{description}
            \item[algebraic process] data and viz transforms are symmetric \cite{kindlmannAlgebraicProcessVisualization2014}
            \begin{equation*}
                \begin{tikzcd}[ampersand replacement=\&]
                    \dtotal \arrow[d, "m"'] \arrow[r, "r_1"] \& \vtotal \arrow[r, "\nu"]  \& \gtotal \arrow[d, "\varphi(m)"] \\
                    \dtotal \arrow[r, "r_2"']                     \& \vtotal \arrow[r, "\nu"'] \& \gtotal                  
                \end{tikzcd}
            \end{equation*}

            \item[language] APT and GoG: syntax, semantics, and grammar\\ 
            \cite{mackinlayAutomatingDesignGraphical1986,mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005}
            \item[functional dependencies] relationship between components \cite{sugibuchiFramwork2009}
            \item[category theory]  \textit{understanding} = \textit{read} $\circ$ \textit{render} \cite{vickersUnderstandingVisualizationFormal2013}

@@

@@

\begin{frame}{Equivariance: Partial Order \& and Stevens' Scales \cite{stevensTheoryScalesMeasurement1946}}
    \begin{figure}
        \includegraphics[width=1\textwidth]{figures/math/monoid_equivariant.png}
    \end{figure}

\begin{table}[H]
    \begin{tabularx}{\textwidth}{|l|l|X|}\toprule
        \textbf{scale} & \textbf{group} & \textbf{constraint} \\\midrule
        nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
        ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
        interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
        ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \bottomrule
    \end{tabularx}
\end{table}
\end{frame}

@@

\begin{frame}{Assembly Function Factory}
\begin{equation}\label{eq:qhat_q_s}
    \vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
\end{equation} 
\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/math/q_hat.png}
\end{figure}
such \gbasepoint\ can be factored out when $\vpreimg = \gbasepoint$
\end{frame}

@@

    def __call__(self, value):
        # convert value to a color
        return [mcolors.to_rgba(self._mapping[v]) for v in values]
\end{minted}
That we can test for action equivariance
\begin{minted}{python}
def test_nominal(values, encoder):
    m1 = list(zip(values, encoder(values)))
    random.shuffle(values)
    m2 = list(zip(values, encoder(values)))
    assert sorted(m1) == sorted(m2)
\end{minted}
\end{frame}
\subsection{Data}
\begin{frame}[fragile]{Fiber Bundle}
\begin{minted}{python}
@dataclass
class FiberBundle:
"""
Attributes
----------
K: {'tables': []}
F: {variable name: type}
"""
    K: dict 
    F: dict
\end{minted}
\end{frame}

@@

\section{References}
\begin{frame}[allowframebreaks]{References}
\printbibliography
\end{frame}
\appendix 
\begin{frame}{Fiber is all possible values a variable can be \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}}
    Given a space of all possible values \ftotal\
    \begin{equation}
        \label{eq:data_types}
        \begin{tikzcd}[ampersand replacement=\&]
            \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] \& \ftotal \arrow[d, "\pi"] \\
            \fnames \arrow[r, "\fsection"']                  \& \ftypes       
        \end{tikzcd}
    \end{equation}
    a fiber component is the restricted space $\ftotal_{\fsection(\fname)}$. 
    \begin{equation}
        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
    \end{equation}
    \begin{description}
        \item[\ftypes] data types of the variables in the dataset 
        \item[\ftotal] disjoint union of all values of type $\ftype \in \ftypes$ 
        \item[\fnames] variable names, $\fname \in \fnames$
        \item[\fttype] \ftotal\ restricted to the data type of a named variable   
    \end{description}
\end{frame}
\begin{frame}{Monoid actions}
    A monoid \monoid\ is a set with
    \begin{description}
        \item[associative binary operator] $\ast:\monoid \times \monoid\rightarrow \monoid$
        \item[identity element] $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. 
    \end{description}
    \begin{block}{left monoid action}
    A set \dfiber\ with an action\cite{nlab:action} $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
        \begin{align*}
            \textbf{associativity}\;& \text{for all } f,g \in \monoid \text{ and } x\in \dfiber,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
            \textbf{identity}\;& \text{for all } x\in \dfiber, e\in \monoid,\,  e\bullet x = x 
        \end{align*}
    \end{block}
\end{frame}
\begin{frame}{Keeping track of sections with sheafs}
    Restriction maps of a sheaf describe how local $\iota^*\dsection$ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
